---
title: "Long term trends in IPD in the US - JAGS models"
author: "Paloma Cárcamo"
date: "2024-08-28"
output: html_document
---

```{r setup}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, rjags, tidybayes)
```

## Load data

```{r}
data_raw <- read_rds("data/ABCs_st_1998_2021.rds")

data <- data_raw |> 
  rename(agec = "Age.Group..years.",
         year = Year,
         st = IPD.Serotype,
         N_IPD = Frequency.Count)  |> 
  mutate(st = if_else(st == '16', '16F', st))  |> 
  group_by(st, year) |> 
  summarize(N_IPD = sum(N_IPD))  |> 
  ungroup()
```

## 1: Only PCV-7 serotypes

### Wrangle data

```{r}
# define  serotypes to include in model
pcv7_st <- c('4','6B','9V','14','18C','19F','23F')

# wrangle dataframe
mod1_df <- data |> 
  complete(st, year, fill = list(N_IPD = 0))  |> 
  arrange(st, year) |> 
  filter(st %in% pcv7_st)  |> 
  group_by(st) |> 
  arrange(year) |> 
  mutate(year_n = row_number()) |> 
  ungroup()

# convert to matrix
mod1_mat <- mod1_df |> 
  reshape2::dcast(year ~ st, value.var = 'N_IPD') |> 
  dplyr::select(-year)
```

### Define model string

```{r}
mod1_string <- "
  model {
    
    for(i in 1:N_years){
      
      for(j in 1:N_sts){
        N_IPD[i,j] ~ dnegbin(prob[i,j], r[j])
        
        prob[i,j]<- r[j] / (r[j] + lambda[i,j])  # Likelihood 
        
        log(lambda[i,j]) <- epsilon1[i,j]        # serotype-specific intercept + AR(1) effect centered around global effects
        }
      }
    
    # Global AR(1) effect
        beta1[1] ~ dnorm(alpha1, (1 - rho_beta^2) * tau.beta1)
        
    for(i in 2:N_years){
      beta1[i] ~ dnorm(alpha1 + rho_beta * beta1[i-1], tau.beta1)
      }
        
    # Serotype-specific AR(1) effect
    
    for(j in 1:N_sts){
      epsilon1[1,j] ~ dnorm(delta1[j] + beta1[1], (1 - rho_eps^2) * tau.eps)
      
      for(i in 2:N_years){
        epsilon1[i,j] ~ dnorm(delta1[j] + beta1[i] + rho_eps * epsilon1[i-1, j], tau.eps)
        }
      }
        
    # Priors
    alpha1 ~ dnorm(0, 1e-4)
    tau_global ~ dgamma(0.01,0.01)

    rho_beta ~ dunif(-1, 1)   # Uniform prior for rho_beta -- global AR(1)
    rho_eps ~ dunif(-1, 1)    # Prior for rho_eps same for all STs
    
    tau.beta1 ~ dgamma(3, 2)  # Tight prior for tau
    tau.eps ~ dgamma(3, 2)    # Tight prior for tau, shared for all serotypes
    
    for(j in 1:N_sts){
       delta1[j] ~ dnorm(0, tau_global)  # serotype means centered around 0
      
       r[j] ~ dunif(0, 250) #serotype dispersion parameter
      }
}"
```

### Specify initial values

```{r}
inits1 = list(".RNG.seed"=c(123), ".RNG.name"='base::Wichmann-Hill')
inits2 = list(".RNG.seed"=c(456), ".RNG.name"='base::Wichmann-Hill')
inits3 = list(".RNG.seed"=c(789), ".RNG.name"='base::Wichmann-Hill')
```

### Create JAGS model object

```{r}
mod1 <- jags.model(textConnection(mod1_string),
                   data = list("N_IPD" = mod1_mat,
                               "N_years" = max(mod1_df$year_n),
                               "N_sts" = ncol(mod1_mat)),
                   inits = list(inits1, inits2, inits3),
                   n.adapt = 10000,
                   n.chains = 3)
```

### Sample posteriors

```{r}
params1 <- c('alpha1', 'epsilon1', 'delta1', 'beta1', 'tau.beta1', 'lambda')

mod1_postsamp <- coda.samples(mod1, params1, n.iter = 10000)
```

### Plot results

```{r}
st_mapping <- setNames(colnames(mod1_mat), 1:length(pcv7_st))
year_mapping <- setNames(unique(mod1_df$year), 1:25)

mod1_summary <- gather_draws(mod1_postsamp, lambda[i,j]) |> 
  median_hdi() |> 
  rename(year = i,
         st = j) |> 
  mutate(st = st_mapping[as.character(st)],
         year = year_mapping[as.character(year)]) |> 
  left_join(mod1_df, by = c("year", "st"))
```

```{r, fig.width = 11}
mod1_summary |> 
  ggplot(aes(x = year, y = .value)) +
  geom_ribbon(aes(x = year, ymin = .lower, ymax = .upper), alpha = 0.2) +
  geom_point(aes(y = N_IPD, color = "Observed")) +
  geom_line(aes(color = "Fitted")) +
  facet_wrap(~st) +
  labs(x = "", y = "# IPD", color = "") +
  theme_minimal()
```

```{r}
# run in console
plot(mod1_postsamp, ask = TRUE)
```

## 2: PCV-7 serotypes plus five uncommon serotypes

```{r}
data |> 
  filter(!(st %in% pcv7_st)) |> 
  group_by(st) |> 
  summarise(N_IPD = sum(N_IPD)) |> 
  arrange(N_IPD)
```

### Wrangle data

```{r}
# define  serotypes to include in model
pcv7plus_st <- c("4","6B","9V","14","18C","19F","23F","15F","33A","35A","9L","29")

# wrangle dataframe
mod2_df <- data |> 
  complete(st, year, fill = list(N_IPD = 0))  |> 
  arrange(st, year) |> 
  filter(st %in% pcv7plus_st)  |> 
  group_by(st) |> 
  arrange(year) |> 
  mutate(year_n = row_number()) |> 
  ungroup()

# convert to matrix
mod2_mat <- mod2_df |> 
  reshape2::dcast(year ~ st, value.var = 'N_IPD') |> 
  dplyr::select(-year)
```

### Create JAGS model object

```{r}
mod2 <- jags.model(textConnection(mod1_string),
                   data = list("N_IPD" = mod2_mat,
                               "N_years" = max(mod2_df$year_n),
                               "N_sts" = ncol(mod2_mat)),
                   inits = list(inits1, inits2, inits3),
                   n.adapt = 10000,
                   n.chains = 3)
```

### Sample posteriors

```{r}
mod2_postsamp <- coda.samples(mod2, params1, n.iter = 10000)
```

### Plot results

```{r}
st_mapping2 <- setNames(colnames(mod2_mat), 1:length(pcv7plus_st))
year_mapping2 <- setNames(unique(mod2_df$year), 1:25)

mod2_summary <- gather_draws(mod2_postsamp, lambda[i,j]) |> 
  median_hdi() |> 
  rename(year = i,
         st = j) |> 
  mutate(st = st_mapping2[as.character(st)],
         year = year_mapping2[as.character(year)]) |> 
  left_join(mod2_df, by = c("year", "st"))
```

```{r, fig.width = 11}
mod2_summary |> 
  ggplot(aes(x = year, y = .value)) +
  geom_ribbon(aes(x = year, ymin = .lower, ymax = .upper), alpha = 0.2) +
  geom_point(aes(y = N_IPD, color = "Observed")) +
  geom_line(aes(color = "Fitted")) +
  facet_wrap(~st) +
  labs(x = "", y = "# IPD", color = "") +
  theme_minimal()
```

```{r}
# run in console
# plot(mod2_postsamp, ask = TRUE)
```

```{r}
# traceplot(mod2_postsamp[, grep("^lambda", varnames(mod2_postsamp))])
```

## 3: Most-common non-PCV-7 serotypes

### Wrangle data

```{r}
# define  serotypes to include in model
nonpcv7_st <- c("3","19A","22F","7F","12F",
                "9N","33F","11A","23A","16F",
                "6C","35B","8","15A","6A",
                "23B","20","31","15B","10A")

# wrangle dataframe
mod3_df <- data |> 
  complete(st, year, fill = list(N_IPD = 0))  |> 
  arrange(st, year) |> 
  filter(st %in% nonpcv7_st)  |> 
  group_by(st) |> 
  arrange(year) |> 
  mutate(year_n = row_number()) |> 
  ungroup()

# convert to matrix
mod3_mat <- mod3_df |> 
  reshape2::dcast(year ~ st, value.var = 'N_IPD') |> 
  dplyr::select(-year)
```

### Create JAGS model object

```{r}
mod3 <- jags.model(textConnection(mod1_string),
                   data = list("N_IPD" = mod3_mat,
                               "N_years" = max(mod3_df$year_n),
                               "N_sts" = ncol(mod3_mat)),
                   inits = list(inits1, inits2, inits3),
                   n.adapt = 10000,
                   n.chains = 3)
```

### Sample posteriors

```{r}
mod3_postsamp <- coda.samples(mod3, params1, n.iter = 10000)
```

### Plot results

```{r}
st_mapping3 <- setNames(colnames(mod3_mat), 1:length(nonpcv7_st))
year_mapping3 <- setNames(unique(mod3_df$year), 1:25)

mod3_summary <- gather_draws(mod3_postsamp, lambda[i,j]) |> 
  median_hdi() |> 
  rename(year = i,
         st = j) |> 
  mutate(st = st_mapping3[as.character(st)],
         year = year_mapping3[as.character(year)]) |> 
  left_join(mod3_df, by = c("year", "st"))
```

```{r, fig.width = 11}
mod3_summary |> 
  ggplot(aes(x = year, y = .value)) +
  geom_ribbon(aes(x = year, ymin = .lower, ymax = .upper), alpha = 0.2) +
  geom_point(aes(y = N_IPD, color = "Observed")) +
  geom_line(aes(color = "Fitted")) +
  facet_wrap(~st) +
  labs(x = "", y = "# IPD", color = "") +
  theme_minimal()
```

```{r}
# run in console
# plot(mod3_postsamp, ask = TRUE)
```

```{r}
# traceplot(mod3_postsamp[, grep("^lambda", varnames(mod3_postsamp))])
```

## 4: Most-common non-PCV-7 serotypes plus five uncommon serotypes

### Wrangle data

```{r}
# define  serotypes to include in model
nonpcv7plus_st <- c("3","19A","22F","7F","12F",
                    "9N","33F","11A","23A","16F",
                    "6C","35B","8","15A","6A",
                    "23B","20","31","15B","10A",
                    "15F","33A","35A","9L","29")

# wrangle dataframe
mod4_df <- data |> 
  complete(st, year, fill = list(N_IPD = 0))  |> 
  arrange(st, year) |> 
  filter(st %in% nonpcv7plus_st)  |> 
  group_by(st) |> 
  arrange(year) |> 
  mutate(year_n = row_number()) |> 
  ungroup()

# convert to matrix
mod4_mat <- mod4_df |> 
  reshape2::dcast(year ~ st, value.var = 'N_IPD') |> 
  dplyr::select(-year)
```

### Create JAGS model object

```{r}
mod4 <- jags.model(textConnection(mod1_string),
                   data = list("N_IPD" = mod4_mat,
                               "N_years" = max(mod4_df$year_n),
                               "N_sts" = ncol(mod4_mat)),
                   inits = list(inits1, inits2, inits3),
                   n.adapt = 10000,
                   n.chains = 3)
```

### Sample posteriors

```{r}
mod4_postsamp <- coda.samples(mod4, params1, n.iter = 10000)
```

### Plot results

```{r}
st_mapping4 <- setNames(colnames(mod4_mat), 1:length(nonpcv7plus_st))
year_mapping4 <- setNames(unique(mod4_df$year), 1:25)

mod4_summary <- gather_draws(mod4_postsamp, lambda[i,j]) |> 
  median_hdi() |> 
  rename(year = i,
         st = j) |> 
  mutate(st = st_mapping4[as.character(st)],
         year = year_mapping4[as.character(year)]) |> 
  left_join(mod4_df, by = c("year", "st"))
```

```{r, fig.width = 11}
mod4_summary |> 
  ggplot(aes(x = year, y = .value)) +
  geom_ribbon(aes(x = year, ymin = .lower, ymax = .upper), alpha = 0.2) +
  geom_point(aes(y = N_IPD, color = "Observed")) +
  geom_line(aes(color = "Fitted")) +
  facet_wrap(~st) +
  labs(x = "", y = "# IPD", color = "") +
  theme_minimal()
```

```{r}
# run in console
# plot(mod4_postsamp, ask = TRUE)
```

```{r}
# traceplot(mod4_postsamp[, grep("^lambda", varnames(mod4_postsamp))])
```